Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions

### 2021-08-03 analysis/paper/shceof.Rmd
revision lean

---
title: "Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation"
titlerunning:  Title
authors: 
- name: Elio Campitelli
  address: |
    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.  
    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.  
    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
  email: elio.campitelli@cima.fcen.uba.ar

@@

For example, @raphael2007 suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies. @irving2015 showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies. Another characteristic of the zonally asymmetric circulation are the Pacific--South American Patterns [@mo2001], which are associated with precipitation anomalies in South America.

Some of these patterns of variability appear to have experienced secular changes. @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.

@@

EOFs are more flexible than Fourier decomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.
The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
@irving2016 characterises the Pacific--South American Pattern in a Fourier framework.
Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
On this new projection, they could identify the PSA using Fourier decomposition.

@@

Our objective is to create an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle. 
We use complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.

We apply this method to characterise the Springtime zonally asymmetric circulation of the extratropical Southern Hemisphere.
To take vertical distribution into account, we analyse both the troposphere and the stratosphere. 
We focus the analysis to the September-October-November because that's the trimester in which tropical forcing of zonal waves is maximised (cita).

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985.
We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle.
We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between `r knitr::combine_words(LatLabel(o3wave_lats))` for each SON and extracting the wave-1 component of the Fourier spectrum.
We chose this latitude band because it is wide enough to capture most of the relevant anomalies.

We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
The Pacific South American patterns (PSA1 and PSA2) were calculated following @mo2001 as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.

Linear trends were computed by Ordinary Least Squares and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom [@wilks2011r].

@@

@@

@@

The phase of principal components is defined up to an additive constant.
For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign.
For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984].
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.

@@

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
This also nearly minimises the correlation with the imaginary part.

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

We performed linear regressions to quantify the association between the cEOFs and other variables.
For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Imaginary and Real parts of each cEOF we fit the equation

@@

where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms respectively.

@@

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:extended-series,).
The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the temporal cEOF.
The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive).

The first cEOF (Fig. \@ref(fig:eof-naive) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
The second cEOF (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is especially apparent at 200 hPa.
There is no apparent phase shift, suggesting a rather barotropic structure.

The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:extended-series,)).
This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
There are no significant simultaneous correlation between the time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability.

There is a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).

This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
A similar observation was made by @raphael2003, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

The regression patterns at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.

@@

Figure \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2. Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.

@@

@@

To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
It is not correlated with the Real cEOF2 at any level.

The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh).

@@

In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.

@@

@@

The relationship between PSAs and the modes described from the cEOFs is also studied.
Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.

The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.

@@

The result obtained here is similar to the one obtained by @irving2016.
They show that the phase distribution of the PSA-like variability they detected is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).

The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.

@@

(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.

@@

The relation between cEOFs and temperature was also evaluated.
Figure \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig. \@ref(fig:eof1-regr-gh)).

@@

The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical))). Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere . This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere. In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979; @wirth1993; @smith1995].

@@

@@

(ref:pp-america-cap) As Figure \@ref(fig:pp-america) but for South America

>>>>

---
title: "Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions"
titlerunning:  Title
authors: 
- name: Elio Campitelli
  address: |
    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.  
    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.  
    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
  email: elio.campitelli@cima.fcen.uba.ar

@@

One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns [@mo2001], which are associated with precipitation anomalies in South America. Besides, @raphael2007 suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies. @irving2015 showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies.

Some of these patterns of variability associated to the zonally asymmetric circulation appear to have experienced secular changes. @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996. \textcolor{red}{¿esto es lo único que encontraste asociado a esta afirmación? en caso de que si, diría de agregar que "ha sido poco estudiado hasta el momento"}

@@

EOFs are more flexible than Fourier decomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.
The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
@irving2016 characterises this pattern in a Fourier framework.
Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle. As a result, they reprojected meridional wind fields so that the path of the PSA laid on the equator.
On this new projection, they could identify the PSA using Fourier decomposition.

@@

Our objective is to create an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle. \textcolor{red}{yo no pondría que el objetivo es crear el índice, sino porfundizar el conocimiento de la variabilidad... utilizando un índice...}
We use complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.

We apply this method to characterise the Springtime zonally asymmetric circulation of the extratropical Southern Hemisphere.
To take vertical distribution into account, we analyse both the troposphere and the stratosphere. 
We focus the analysis to the September-October-November because that's the trimester in which tropical forcing of zonal waves is maximised (cita).
\textcolor{red}{Este parráfo me parece que quedo un poco colgado. Creo que sería importante que esté desde más arriba la idea de la primavera, ya que todo lo otro es muy general, pero al final el análisis solo se va a restringir a esta estación}

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 reanalysis [@era5] for the period 1979-2019.
Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985.
We used Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) each variable between `r knitr::combine_words(LatLabel(o3wave_lats))` for each SON and extracting the wave-1 component of the Fourier spectrum.
We chose this latitude band because it is wide enough to capture most of the relevant anomalies for southern hemisphere mid-latitudes.

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according to @baldwin2009.
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
The Pacific South American patterns (PSA1 and PSA2) were calculated following @mo2001 as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons \textcolor{red}{leyendo esto me queda la duda. Vos calculaste esto usando todos los meses y luego tomaste el valor de SON, no? Creo que valdría la pena aclararlo mejor, por lo menos a mi me resulta confuso}.

Linear trends were computed by Ordinary Least Squares (OLS) and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom [@wilks2011r].

@@

\textcolor{red}{En la figura 1 te quedaron muy chicos los subplots, amplialos un poco}.

@@

\textcolor{red}{Los colores de la Tabla 1 no dicen mucho en la medida que no se conoce cuál sería la paleta. Quizás se pueden remarcar en negrita ambos aclarando que son significativos (lo son?)}.

@@

The phase of principal components is defined up to an additive constant.
For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign.
For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984].
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase. For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
This also nearly minimises the correlation with the imaginary part. For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

@@

We performed linear regressions to quantify the association between the cEOFs and other variables.
For each cEOF, we create regression maps by fitting a multiple linear model involving both the real and the imaginary part.
To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Imaginary and Real parts of each cEOF we fit the equation

@@

where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients for Imaginary and Real parts respectively, $X_0$ and $\epsilon$ are the constant and error terms respectively.

@@

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:extended-series).
The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the temporal cEOF.
The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive).

\textcolor{red}{tengo la impresión de que en la figura 2 podría estar un poco más grandes los subplots para que se vean más fácilmente los detalles}

The first cEOF (Fig. \@ref(fig:ceofs-1) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
The cEOF2 (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is especially apparent at 200 hPa.
There is no apparent phase shift, suggesting a rather barotropic structure.

\textcolor{red}{se me vino una duda, ¿se pueden comparar de algún modo las amplitudes que mostras en esta figura? en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera}

The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:extended-series)). \textcolor{red}{Esto vale también para el cEOF2, ¿no? en ese caso, lo diría más en gral}
This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
There are no significant simultaneous correlation between the time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability.

\textcolor{red}{En la figura 3 agregaría dos cosas: i)el eje y en la columna derecha (si la queres ver más grande perdes la referencia) ii) creo que serviría agregar una líneas verticales en la grilla para poder tener una idea un poco mejor de a qué años corresponde más o menos cada valor. pueden ser un poco más finas que las que ya están, pero creo que sería una referencia util para visualmente poder decir a qué año corresponde cada punto. Además, fijate que la figura te quedo de un ancho mayor que el de la página y te quedo un "EOF" en el eje-y a la izquierda que no creo que sea lo más correcto.}

There is only a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown). This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
A similar observation was made by @raphael2003, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.
\textcolor{red}{Creo que acá faltaría decir que el modo del trabajo de Raphael se parece al que estás mostrando vos, sino no tendria mucho sentido la comapración.}

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

\textcolor{red}{La figura 4 la veo muy saturada con los puntos que indican la significancia. ¿se podrán hacer con un tamaño menor? ¿Las unidades son correctas? Me llama la atención que te puedan dar tantos cientos de metros de altura geopotencial.}

The regression patterns for cEOF1 at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
It is only possible to partially distinguish the wave 1 pattern in relation with the real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.

@@

Figure \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2. Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.

\textcolor{red}{Figura 5. Me resulto raro que vaya hasta 999 en vez de 1000 }

@@

\textcolor{red}{Este título queda raro porque después hablas de modos de variabilidad climática y no de lo que llamaríamos variables}

@@

To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
The SAM index is statistically signficant correlated  with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
Correlations between SAM and the Real cEOF2 are non-significant.

\textcolor{red}{Fig. 6. Te quedan muy pegados los subplots de la columna izquierda con los de la derecha. Recomendaría aumentar el espacio intermedio y repetir el eje y en las figuras de la derecha. Además podrías agregar el y-label (Level) a la izquierda.}

The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by \textcolor{red}{Campitelli et al 2021}.

@@

In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
The Imaginary cEOF2 explains up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of A-SAM variance, reached the maximum value at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2). 
Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.

@@

\textcolor{red}{Tabla 2. Mismo comentario que antes con lo del background color (no dice mucho si no se conoce la escala), quizás podes destacar en negritas según la significancia. ¿Está bien el intervalo de confianza de PSA2 vs cEOF2 Im? Hay algo raro ahí}

@@

The relationship between PSAs and the modes described from the cEOFs is also studied.
Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
\textcolor{red}{Valdría la pena aclarar por qué no consideras el cEOF1 para esta discusión}
As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.

The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO. \textcolor{red}{Ojo que todavía no estuvo la discusión con lo del ENSO, así que esto lo deberías decir más adelante.}
Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.

@@

The result obtained here is similar to the one obtained by @irving2016.
They show that the phase distribution of the PSA-like variability obtained by them is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).

The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillations.
As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.

@@

(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature \textcolor{red}{Unidades}.

@@

The relation between cEOFs and air temperature was also evaluated.
Figure \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig. \@ref(fig:eof1-regr-gh)).

@@

The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical)). Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere. This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere. In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone. On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in adittion 90º out of phase with temperature anomalies [@hartmann1979; @wirth1993; @smith1995].

@@

\textcolor{red}{Revisa lo mismo que para la otra figura del tipo. }

@@

(ref:pp-america-cap) As Figure \@ref(fig:pp-oceania) but for South America
